Supplement

### 2022-06-15 analysis/article.Rmd
feat: further work on the manuscript

tar_load(ameldcfg)
```

@@

Where $LL$ and $UL$ are the lower and upper reference limit of
the corresponding laboratory diagnostic, respectively.
Missing laboratory measurements were replaced with a *zlog* value of zero.
In similar manner if information about complications or
comorbidities were missing we treat them as not present.

@@

The final model was compared against the MELD, MELD-Na and MELD-Plus7 score
using the area under the time-dependent ROC curve (AUC)
based on the nonparametric inverse probability of censoring weighting estimator
(IPCW) [@blanche2013, @R-timeROC].

@@

```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms.", echo = FALSE}
r <- bmrk_results$clone()
id <- grep("^scale", x = r$learners$learner_id, invert = TRUE, value = TRUE)
r$filter(task_id = "zlog_eldd", learner_id = id)
autoplot(r) +
    geom_boxplot(aes(fill = learner_id)) +
    geom_jitter(position = position_jitter(0.2)) +
    scale_fill_viridis(discrete = TRUE) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

TODO: mention best ml algorithm (\@ref(fig:benchmark-plot))

@@

@@

- more variables, harder to cheat
- comparison of different predictors of MELD-Plus7 (+MELD-Plus9) vs our features

Sodium is not part of our model.
In line with the results presented in @kartoun2017 we could show that
adding sodium to MELD the discrimination performance increases just
insignificantly.

It has shown that women get lower scores in MELD-Na and therefore
are less likely getting an urgent listing for liver transplantation
[@Gaertner2022].
To address sexual differences in laboratory measurements we applied the *zlog*
transformation [@hoffmann2017].
Additionally this transformation yield comparable laboratory measurements
across different clinical laboratories and methods which could improve the
objectivity of the allocation process.
The already nearly normal scaled *zlog*-transformed laboratory measurements
allow us to avoid the sample depending scaling prior to model development.
Both should increase the external applicability of our approach.

Additionally to laboratory data we investigated
`r (ncol(raw_data) - length(cn_labs) - 2)`
clinical data but none of them were chosen in our model selection process.
@kartoun2017 used a similar penalizing approach and neither select any clinical
data. Especially in retrospective analysis the clinical data are often not
detailed enough.

@@

@@

Our study has limitations. Principally it is a single center, retrospective
analysis and without an external validation cohort.
The data were collected during the evaluation process for liver transplantation
at the University Hospital of Leipzig.
There was no controlled follow-up process. Instead the follow-up information
were recored at the next clinical admission.
Furthermore we excluded all patients who had already a liver transplantation
which could artificial lower the disease severity of the cohort.

@@

>>>>

tar_load(ameldcfg)
tar_load(rngr)
```

@@

Where $LL$ and $UL$ are the lower and upper reference limit of
the corresponding laboratory diagnostic, respectively.
If more than 10 % of laboratory measurements are missing the variable were
excluded from further analysis.
The remainder of missing laboratory measurements were replaced with
a *zlog* value of zero.
In similar manner if information about complications or
comorbidities were missing we treat them as not present.

@@

The final model was compared against the MELD, MELD-Na and MELD-Plus7 score
using the area under the time-dependent ROC curve (AUC)
based on the nonparametric inverse probability of censoring weighting estimator
(IPCW) [@blanche2013, @R-timeROC]. Testing was applied as described in @blanche2013. A $p < 0.5$ was considered as statistical significant difference.

@@

## RCV model

The penalized regression and random forest models
result in a similar high median performance in the benchmark process.
In favour of simpler models we chose a lasso based approach for the final model
development.
The classical cox model, the data and computational intense algorithms
extreme gradient boosting, support vector machines and neural networks yield a
worse performance.

```{r coeftbl, echo = FALSE}
tar_load(bootrcv)
cf <- coef(bootrcv$fit)
cf <- setNames(cf@x, rownames(cf)[cf@i + 1])
o <- order(-abs(cf))
cfs <- cf[o]
cfnms <- lapply(
    labelled_meld_data[names(cfs)],
    function(nm)sub(" *\\[.*$", "", attr(nm, "label"))
)
cfnms[!lengths(cfnms)] <- names(cfnms)[!lengths(cfnms)]
knitr::kable(
    setNames(cfs, cfnms),
    col.names = "Coefficient",
    caption = "Model coefficients",
    digits = 3
)
```

## Model comparison

```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE}
library("timeROC")
tar_load(timeROC_MELD)
tar_load(timeROC_MELDNa)
tar_load(timeROC_MELDPlus7)
tar_load(timeROC_RCV)

col <- viridisLite::viridis(5)
m <- list(
    RCV = timeROC_RCV,
    MELD = timeROC_MELD,
    "MELD-Na" = timeROC_MELDNa,
    "MELD-Plus7" = timeROC_MELDPlus7
)
invisible(plot_surv_roc(m, timepoint = 90, col = col[-2]))
```

```{r roctest}
tp <- timeROC::compare(
    timeROC_RCV, timeROC_MELDNa,
    adjusted = TRUE
)$p_values_AUC
rownames(tp) <- paste(
    c("Non-adjusted", "Adjusted"), "$p$-values"
)
knitr::kable(
    tp,
    caption = paste0(
        "Test comparison between AUC for MELD-Na and RCV model ",
        "according to @blanche2013 at day ",
        paste0(timeROC_RCV$times, collapse = ", "), ".",
        "$p$-values are shown as non-adjusted and adjusted for ",
        "multiple testing."
    ),
    digits = 2
)
```

```{r bootrcv, fig.width = 7, fig.height = 7, echo = FALSE}
tar_load(ameldcfg)
tar_load(amelddata)
tar_load(zlog_data)
par(cex = 0.9)
plot(bootrcv, what = "calibration")

ps <- lapply(
    zlog_data[paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7"))],
    function(p) {
        ctpnts <- cutpoints(p, n = ameldcfg$m)
        f <- cut(p, ctpnts, include.lowest = TRUE)
        list(
            predicted = groupmean(p, f = f),
            observed = observed_survival(
                amelddata$y, f = f, times = ameldcfg$times
            )
        )
    }
)
names(ps) <- c("MELD", "MELD-Na", "MELD-Plus7")

for (i in seq_along(ps)) {
    lines(
        ps[[i]]$predicted, ps[[i]]$observed, col = col[i + 2],
        type = "b", pch = 19
    )
}
legend("topleft", col = col[3:5], legend = names(ps), pch = 19, bty = "n")
```

```{r basehaz, echo = FALSE}
bh <- basehaz(
    bootrcv$fit, x = amelddata$x, y = amelddata$y, s = "lambda.1se",
    times = timeROC_RCV$times, centered = FALSE
)
s0 <- setNames(bh$hazard, bh$time)
k <- knitr::kable(
    t(s0),
    digits = 4,
    caption = paste0(
        "Underlying survival function; S~0~: baseline survival estimate"
    )
)
k[3] <- paste("|t [days]", k[3])
k[4] <- paste0("|:------", k[4])
k[5] <- paste("|S~0~ (t)", k[5])
k
```

The 90-day survival estimate of a given patient could be calculated as shown in
\@ref(eq:survcalc).

```{r survivalcalc, results = "asis"}
cfnmsltx <- paste0("\\text{", gsub(" ", "\\ ", cfnms), "}")
cat(
    "$$\\begin{align}",
    "S &= S_{0,90}^{\\exp{LP}} \\\\",
    "&= ", round(s0["90"], 4), "^{\\exp{(",
        paste0(
            round(cfs[1], 4), " * zlog(", cfnmsltx[1], ")",
            paste0(
                ifelse(sign(cfs[-1]) > 0, "+", ""),
                round(cfs[-1], 4), " * zlog(", cfnmsltx[-1], ")",
                collapse = ""
            ),
            collapse = ""
        ),
    ")}}",
    #"(\\#eq:survcalc)",
    "\\end{align}$$"
)
```

Where $S_{0}$ is the baseline survival estimate from
\@ref(tab:basehaz) and $LP$ is the linear predictor of the coefficients
in \@ref(tab:coeftbl).

@@

Although the Organ Procurement and Transplantation Network recommends the use
of MELD-Na over MELD the sodium does not add any information to the score.
Neither in the bootstrap selection nor in the variable importance of a
random forest model the sodium gains any important weight
(in bootstrap selection rank
`r as.vector(rank(-ameld:::.selvars(bootrcv$models, bootrcv$s))["NA_S"])`
of
`r ncol(amelddata$x)` and in random forest variable importance
`r as.vector(rank(-rngr$variable.importance)["NA_S"])` of
`r ncol(amelddata$x)`,
see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
Already @kartoun2017 have shown that the
discrimination performance of MELD-Na is insignificantly different from the
classical MELD.
Consequentially it is not a component of our model.

It has shown that women get lower scores in MELD-Na and therefore
are less likely getting an urgent listing for liver transplantation
[@gaertner2022].
To address sexual differences in laboratory measurements we applied the *zlog*
transformation [@hoffmann2017].
Additionally this transformation yield comparable laboratory measurements
across different clinical laboratories and methods.
It also reduce the influence of age- and sex-related differences,
which could improve the objectivity of the allocation process.
The already nearly normal scaled *zlog*-transformed laboratory measurements
allow us to avoid the sample depending scaling prior to model development.
Both should increase the external applicability of our approach.

@@

- comparison of different predictors of MELD/MELD-Plus7 (+MELD-Plus9) vs our features

Results from machine learning can aid in finding clinical meaningful patterns
and guide further research.
For example, the systemic inflammatory response syndrome is an
independent risk factor for poor outcome in patients with cirrhosis
[@thabut2007].
However neither MELD nor MELD-Na taking inflammation into account.
In contrast IL6 is the most important variable in our model. Even if IL6 is
excluded the algorithms select CRP as one of the most important variables which
highlights the important role of inflammation in the prognosis of patients
with end-stage liver disease.
We and other researchers have already shown that adding inflammatory markers,
namely IL6 or CRP, to the classical MELD score increase the accuracy of the
90-day mortality prediction [@remmler2018; @cervoni2016].

Beside laboratory measurements we investigated
`r (ncol(raw_data) - length(cn_labs) - 2)`
clinical data but none of them were chosen in our model selection process.
@kartoun2017 used a similar penalizing approach and neither select any clinical
data. Especially in retrospective analysis the clinical data are often not
detailed enough to keep valuable information.

@@

<!--
As we shown above machine learning can aid in finding clinical meaningful
patterns.
-->

@@

Our study has several limitations.
Principally it is a single center, retrospective analysis and without an
external validation cohort.
The data were collected during the evaluation process for liver transplantation
at the University Hospital of Leipzig.
There was no controlled follow-up process. Instead the follow-up information
were recored at the next clinical admission.
Furthermore we excluded all patients who had already a liver transplantation
which could artificial lower the disease severity of the cohort.
Next most of our patients suffered from ethyltoxic liver cirrhosis
(`r .tbl1("Ethyltoxic", pattern = "{p}%")`) which may
limited the transfer to other etiologies.

@@

# Data and software availability

The complete code and reproducible analysis could be found at
https://ampel-leipzig.github.io/ampel-leipzig-meld/.
Additionally all data are available in the `ameld` R package at
https://github.com/ampel-leipzig/ameld for further investigations.

### 2022-06-15 analysis/supplement.Rmd
refactor: move common yaml headers into _site.yml
feat: add first discussion draft
feat: further work on the manuscript

---
title: "Supplement"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r supplement_setup, message = FALSE}
library("targets")
```

>>>>

---
title: "Supplement"
output:
  bookdown::html_document2:
    toc: false
---

```{r knitr_setup, include=FALSE}
```

```{r supplement_setup, message = FALSE}
library("ameld")
library("data.table")
library("targets")
library("mlr3")
library("mlr3misc")
library("mlr3proba")
library("mlr3viz")
library("viridis")
library("ggplot2")
```

# Benchmark results of machine learning algorithms

```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms.", echo = FALSE, fig.width = 18, fig.height = 12}
tar_load(bmrk_results)
r <- bmrk_results$clone()
id <- grep("^scale", x = r$learners$learner_id, invert = TRUE, value = TRUE)
r$filter(task_id = "zlog_eldd", learner_id = id)
autoplot(r) +
    geom_boxplot(aes(fill = learner_id)) +
    geom_jitter(position = position_jitter(0.2)) +
    scale_fill_viridis(discrete = TRUE) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

```{r benchmark-table}
agg <- bmrk_results$aggregate(msr("surv.cindex", id = "harrell"))
agg[, `:=`
    (nr = NULL, resample_result = NULL, resampling_id = NULL, iters = NULL)
]
agg <- subset(agg, task_id == "zlog_eldd" & !grepl("^scale", learner_id))
agg[, task_id := NULL]
setorder(agg, -harrell)
knitr::kable(agg, col.names = c("Learner ID", "Harrell C"))
```

# AUROC trend

```{r roc-plot, fig.width = 18, fig.height = 12}
tar_load(timeROC_MELD)
tar_load(timeROC_MELDNa)
tar_load(timeROC_MELDPlus7)
tar_load(timeROC_RCV)

m <- list(
    MELD = timeROC_MELD,
    "MELD-Na" = timeROC_MELDNa,
    "MELD-Plus7" = timeROC_MELDPlus7,
    RCV = timeROC_RCV
)
plot_surv_roc_trend(m)
```

# Variable importance

```{r vimpbs, fig.width = 8, fig.height = 10, fig.cap = "Variable importance by frequency of bootstrap selections.", echo = FALSE, message = FALSE}
tar_load(bootrcv)
plot(bootrcv, what = "selected", cex = 0.5)
```

```{r vimprf, fig.width = 8, fig.height = 10, fig.cap = "Variable importance by logrank in random forest.", echo = FALSE, message = FALSE}
tar_load(rngr)
plot_dots(
    sort(rngr$variable.importance),
    main = "Variable importance (random forest)",
    xlab = rngr$splitrule
)
```
